Methods for Analyzing Models of\\ Rod-Shaped Bacteria

### 2025-09-25 paper/main.tex
link to correct subsection of supplement

We will use our previously developed computational model and apply it directly to the data provided
by the experiments of Amir et. al~\cite{Amir2014,Amir2014_2}.
These images show a singular \textit{E.Coli} which is fixed at the bottom and is subject to a
horizontal fluid flow that exerts a force onto the rod.
This setup is displayed in figure~\ref{fig:amir-bending-simulation} (A).
In order to model this setup, we also fixate the lower-most vertex of our rod at the bottom of the
simulation domain.
Furthermore, we introduce a force $\vec{F}_i = C |\vec{x}_i - \vec{x}_{i+1}|\sin(\alpha_i)\vec{e}_x$
that acts similarlyl to Stoke's Law \textbf{TODO Citation} onto the normal component of each
rod-segment where $\alpha_i$ is the angle of the rod edge between $\vec{x}_i$ and $\vec{x}_{i+1}$
and the x-axis.
We extract the position of the rod from the provided image and use it as intial value.
Our simulation is split into two phases, one with an external force that initially bends the rod,
and a second one during which the force is disabled and the cell returns to its original shape.
We compare the simulated positions with the extracted ones and calculate the euclidean distance
between them.
This value serves as the cost function that allows us to estimate the parameters of this model.
Since this model is void of any interactions or cell division, we are left with 5 parameters:
Magnitude of the drag force $C$, growth rate $\alpha$, rigidity of the rod $\eta$, spring tension
$\gamma$ and damping constant $\lambda$.
Figure~\ref{fig:amir-bending-simulation} (B) shows a snapshot of the initial (orange) and final
(blue) position of the rod using the optimized parameters.
While the dragging force, growth rate and damping can be identified properly, the same
can not be said for the rod rigidity and spring tension which both exhibit non-identifiabilities.
However this behaviour is expected since, the shape of the rod is almost exclusively straight,
thus adding no further information except that a lower threshold of rigidity and spring tension
should be present.
Note that each point represented in the subfigures~\ref{fig:amir-bending-simulation} (C-G) was
optimized again, meaning that higher values of rigidity and spring tension can also lead to higher
values of drag forces.
Furthermore the used images only contain snapshots of a fully-bent or fully-relaxed rod without any
intermediate steps that cannot fully capture its dynamic.
This effectively limits us to 2 datapoints and is likely the key factor that leads to the two local
minima observed in the damping profile.
This is further supported by the observation that the contributions to the cost function is evenly
distributed between the bent and relaxed position of our system.
In order to better understand this system, we reduce it by inserting a new parameter that determines
the values of rod rigidity and spring tension via $\gamma=\gamma_0/s$ and $\eta=\eta_0 s$.
We then optimize for this parameter and pick $\gamma_0=\SI{15}{\pixel}$ with
$\eta_0=\SI{150}{\pixel\per\second\squared}$ which corresponds to $60\%$ of the previously considered
interval.
With this change, all parameters are identifiable.
The corresponding plots can be seen in \ref{sec:supplement-parameter-profiles}.

\begin{figure}
    \centering
    \begin{minipage}{0.33\textwidth}
        \begin{tikzonimage}[width=\textwidth]
            {figures/amir-elastic-000032-wider.png}%
            \node at (0.01, 0.01+17pt)[anchor=north west, rectangle, draw, white, minimum width=15pt, minimum height=15pt]{\textbf{A}};
        \end{tikzonimage}%
    \end{minipage}%
    \begin{minipage}{0.66\textwidth}
        \begin{tikzonimage}[width=0.33\textwidth]
            {figures/crm_amir/profiles-full/fit-comparison.pdf}%
            \node at (0.15, 0.86)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{B}};
        \end{tikzonimage}%
        \begin{tikzonimage}[width=0.33\textwidth]
            {figures/crm_amir/profiles-full/drag force.pdf}%
            \node at (0.15, 0.86)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{C}};
        \end{tikzonimage}%
        \begin{tikzonimage}[width=0.33\textwidth]
            {figures/crm_amir/profiles-full/growth rate.pdf}%
            \node at (0.15, 0.86)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{D}};
        \end{tikzonimage}\\
        \begin{tikzonimage}[width=0.33\textwidth]
            {figures/crm_amir/profiles-full/rod rigidity.pdf}%
            \node at (0.15, 0.86)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{E}};
        \end{tikzonimage}%
        \begin{tikzonimage}[width=0.33\textwidth]
            {figures/crm_amir/profiles-full/spring tension.pdf}%
            \node at (0.15, 0.86)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{F}};
        \end{tikzonimage}%
        \begin{tikzonimage}[width=0.33\textwidth]
            {figures/crm_amir/profiles-full/damping.pdf}%
            \node at (0.15, 0.86)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{G}};
        \end{tikzonimage}%
    \end{minipage}
    \caption{
        TODO \cite{Amir2014_2}
    }
    \label{fig:amir-bending-simulation}
\end{figure}

@@

%###################################################################################################
\subsection{Parameter Estimation Results (TODO)}
\label{sec:supplement-parameter-profiles}
%TODO {explain profiles}
\begin{figure}
    \centering
    \includegraphics[width=0.33\textwidth]
        {figures/crm_fit/profiles/growth-rate-0.pdf}%
    \includegraphics[width=0.33\textwidth]
        {figures/crm_fit/profiles/growth-rate-1.pdf}%
    \includegraphics[width=0.33\textwidth]
        {figures/crm_fit/profiles/growth-rate-2.pdf}\\%
    \includegraphics[width=0.33\textwidth]
        {figures/crm_fit/profiles/growth-rate-3.pdf}%
    \includegraphics[width=0.33\textwidth]
        {figures/crm_fit/profiles/growth-rate-4.pdf}%
    \includegraphics[width=0.33\textwidth]
        {figures/crm_fit/profiles/growth-rate-5.pdf}
    \caption{Parameter estimation of growth rates for individual cells}%TODO extend caption}
    \label{fig:parameter-estimates-supplement-growth-rates}
\end{figure}

>>>>

We will use our previously developed computational model and apply it directly to the data provided
by the experiments of Amir et. al~\cite{Amir2014,Amir2014_2}.
These images show a singular \textit{E.Coli} which is fixed at the bottom and is subject to a
horizontal fluid flow that exerts a force onto the rod.
This setup is displayed in figure~\ref{fig:amir-bending-simulation} (A).
In order to model this setup, we also fixate the lower-most vertex of our rod at the bottom of the
simulation domain.
Furthermore, we introduce a force $\vec{F}_i = C |\vec{x}_i - \vec{x}_{i+1}|\sin(\alpha_i)\vec{e}_x$
that acts similarlyl to Stoke's Law \textbf{TODO Citation} onto the normal component of each
rod-segment where $\alpha_i$ is the angle of the rod edge between $\vec{x}_i$ and $\vec{x}_{i+1}$
and the x-axis.
We extract the position of the rod from the provided image and use it as intial value.
Our simulation is split into two phases, one with an external force that initially bends the rod,
and a second one during which the force is disabled and the cell returns to its original shape.
We compare the simulated positions with the extracted ones and calculate the euclidean distance
between them.
This value serves as the cost function that allows us to estimate the parameters of this model.
Since this model is void of any interactions or cell division, we are left with 5 parameters:
Magnitude of the drag force $C$, growth rate $\alpha$, rigidity of the rod $\eta$, spring tension
$\gamma$ and damping constant $\lambda$.
Figure~\ref{fig:amir-bending-simulation} (B) shows a snapshot of the initial (orange) and final
(blue) position of the rod using the optimized parameters.
While the dragging force, growth rate and damping can be identified properly, the same
can not be said for the rod rigidity and spring tension which both exhibit non-identifiabilities.
However this behaviour is expected since, the shape of the rod is almost exclusively straight,
thus adding no further information except that a lower threshold of rigidity and spring tension
should be present.
Note that each point represented in the subfigures~\ref{fig:amir-bending-simulation} (C-G) was
optimized again, meaning that higher values of rigidity and spring tension can also lead to higher
values of drag forces.
Furthermore the used images only contain snapshots of a fully-bent or fully-relaxed rod without any
intermediate steps that cannot fully capture its dynamic.
This effectively limits us to 2 datapoints and is likely the key factor that leads to the two local
minima observed in the damping profile.
This is further supported by the observation that the contributions to the cost function is evenly
distributed between the bent and relaxed position of our system.
In order to better understand this system, we reduce it by inserting a new parameter that determines
the values of rod rigidity and spring tension via $\gamma=\gamma_0/s$ and $\eta=\eta_0 s$.
We then optimize for this parameter and pick $\gamma_0=\SI{15}{\pixel}$ with
$\eta_0=\SI{150}{\pixel\per\second\squared}$ which corresponds to $60\%$ of the previously considered
interval.
With this change, all parameters are identifiable.
The corresponding plots can be seen in \ref{subsec:supplement-parameter-profiles-amir}.

\begin{figure}
    \centering
    \begin{minipage}{0.33\textwidth}
        \begin{tikzonimage}[width=\textwidth]
            {figures/amir-elastic-000032-wider.png}%
            \node at (0.01, 0.01+17pt)[anchor=north west, rectangle, draw, white, minimum width=15pt, minimum height=15pt]{\textbf{A}};
        \end{tikzonimage}%
    \end{minipage}%
    \begin{minipage}{0.66\textwidth}
        \begin{tikzonimage}[width=0.33\textwidth]
            {figures/crm_amir/profiles-full/fit-comparison.pdf}%
            \node at (0.15, 0.86)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{B}};
        \end{tikzonimage}%
        \begin{tikzonimage}[width=0.33\textwidth]
            {figures/crm_amir/profiles-full/drag force.pdf}%
            \node at (0.15, 0.86)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{C}};
        \end{tikzonimage}%
        \begin{tikzonimage}[width=0.33\textwidth]
            {figures/crm_amir/profiles-full/growth rate.pdf}%
            \node at (0.15, 0.86)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{D}};
        \end{tikzonimage}\\
        \begin{tikzonimage}[width=0.33\textwidth]
            {figures/crm_amir/profiles-full/rod rigidity.pdf}%
            \node at (0.15, 0.86)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{E}};
        \end{tikzonimage}%
        \begin{tikzonimage}[width=0.33\textwidth]
            {figures/crm_amir/profiles-full/spring tension.pdf}%
            \node at (0.15, 0.86)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{F}};
        \end{tikzonimage}%
        \begin{tikzonimage}[width=0.33\textwidth]
            {figures/crm_amir/profiles-full/damping.pdf}%
            \node at (0.15, 0.86)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{G}};
        \end{tikzonimage}%
    \end{minipage}
    \caption{
        Subfigure (A) shows a snapshot of the experiments performed by \cite{Amir2014_2}.
        Division of the bacteria is inhibited and they are trapped within a mother machine such that
        a controlled flow can be exerted on them.
        (B) A comparison of the initial and bent state show good visual agreement of our prediction
        with the experimental data.
        Subfigures (C-G) show likelihood profiles for the individual parameters.
    }
    \label{fig:amir-bending-simulation}
\end{figure}

@@

%###################################################################################################
\subsection{Parameter Estimation Results (TODO)}
\label{sec:supplement-parameter-profiles}

\subsubsection{Amir}
\label{subsec:supplement-parameter-profiles-amir}

The parameter ranges were chosen such that a suitable 
Plots are chosen for visual ease; parameter ranges have been widened to ensure results are
accurate (maybe just leave this in the appendix).

%TODO {explain profiles}
\begin{figure}
    \centering
    \includegraphics[width=0.33\textwidth]
        {figures/crm_fit/profiles/growth-rate-0.pdf}%
    \includegraphics[width=0.33\textwidth]
        {figures/crm_fit/profiles/growth-rate-1.pdf}%
    \includegraphics[width=0.33\textwidth]
        {figures/crm_fit/profiles/growth-rate-2.pdf}\\%
    \includegraphics[width=0.33\textwidth]
        {figures/crm_fit/profiles/growth-rate-3.pdf}%
    \includegraphics[width=0.33\textwidth]
        {figures/crm_fit/profiles/growth-rate-4.pdf}%
    \includegraphics[width=0.33\textwidth]
        {figures/crm_fit/profiles/growth-rate-5.pdf}
    \caption{Parameter estimation of growth rates for individual cells}%TODO extend caption}
    \label{fig:parameter-estimates-supplement-growth-rates}
\end{figure}
